SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-07 back.tex
finished a pass

Kim {\em et. al.} proposed Signature Path Prefetcher (SPP)~\cite{SPP},
a confidence-based lookahead prefetcher.  SPP creates a signature
associated with a page address by compressing the history of
accesses. By correlating the signature with future likely delta
patterns, SPP learns both simple and complicated memory access
patterns quickly.  While the basic idea of perceptron based prefetch
filtering is applicable to any look-ahead prefetcher, we develop a
practical implementation of the idea
using SPP as our baseline. Here we describe the basic architecture of SPP.

\noindent \textbf{Signature Table:} As shown in figure
\ref{fig:spp_structure}, the Signature Table keeps track of 256 most
recently accessed pages. It is meant to capture memory access patterns
within a page boundary. SPP indexes into an entry of the Signature
Table using the page number. For each entry corresponding to a page,
the table stores a `last block offset' and an `old signature'. Last
block offset is the block offset of the last memory access of that
given page. The block offset is calculated with respect to the page
boundary. The signature is a 12-bit compressed representation of the
past few memory accesses for that page. The signature is calculated
as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\;
(\,Delta\,)$$ Delta is the numerical difference between the block
offset of the current and the previous memory access. In case a
matching page entry is found, the stored signature retrieved and used
to index into the Pattern
Table. This process is illustrated by figure \ref{fig:spp_update}.

\noindent \textbf{Pattern Table:} The Pattern Table is indexed by the
signature generated from the Signature Table. Pattern Table holds
predicted delta patterns and their confidence estimates. Each entry
indexed by the signature holds up to 4 unique delta predictions.

@@

%% NOTE: Can we do away with this example? 
%% If it looks like we are writing too much about SPP
%% EDIT: Decided to reorganize and remove this for now.
%% 	Consider the case that incoming page number 10 with a block offset 3
%% 	finds a match in the ST. The retrieved pattern signature is 0x30 and
%% 	the Last Offset is 1. Since now there is the Last Block Offset (1),
%% 	Incoming Block Offset (3), Old Signature (0x30), and New Signature
%% 	calculated as per above equation (0x182), SPP can infer
%% 	non-speculatively that the given pattern of memory accesses (as
%% 	captured in the Old Signature) leads to the particular Delta. In
%% 	general, Delta is defined as the difference between the prefetch
%% 	suggestion block and the initial block which triggered the prefetch.
%% 	In this case, since SPP is in learning phase, it is defined as the
%% 	difference between the Incoming Block Offset and the Last Block Offset
%% 	(+2 in this case). This, in turn generates the new memory pattern
%% 	(New Signature). This newly learned signature pattern and the delta
%% 	is stored in the Pattern Table.
\noindent \textbf{Prefetch Table:} The Prefetch Table is a 1024 entry
1-way associative table that keeps a record of last few entries
prefetched. This proves useful for updating the prefetcher states if a
tracked prefetch leads to a demand hit or a cache
eviction. \textit{Note:} The original paper refers to this as the
Prefetch Filter but since in our context filter refers to the
perceptron filter, we will be calling this structure as the Prefetch
Table to avoid any confusion.

\noindent \textbf{Lookahead Prefetching:} On each trigger, SPP goes
down the program speculation path using its own prefetch
suggestion. Using current prefetch as the baseline, it re-accesses the
Pattern Table to generate further prefetches.  It repeats the cycle of
accessing Pattern Table and updating the signature based on highest
confidence prefetch from the last iteration. The iteration counter on
which SPP manages to predict prefetch entries in the lookahead manner
is characterized as its `depth'. While doing so, SPP also keeps
compounding the confidence in each depth. Thus as depth increases,
overall confidence keeps decreasing.

\noindent \textbf{Confidence Tracking}: The Pattern Table keeps a
count of hits to each signature through a counter
C\textsubscript{sig}. The number of hits for a given delta per
signature are tracked using a counter C\textsubscript{delta}.  The
confidence for a given delta is approximated through
C\textsubscript{d} = C\textsubscript{delta} /
C\textsubscript{sig}. When SPP enters into a lookahead mode, the path
confidence P\textsubscript{d} is given as:
$$P\textsubscript{d} \;=\; \alpha  \;.\;  C\textsubscript{d}  \;.\;
P\textsubscript{d-1}$$ Here $\alpha$ represents the global accuracy,
calculated as the ratio of the number of prefetches which led to a demand hit
to the number of prefetches recommended in total. The range of $\alpha$ is [0,1].
$D$ is the lookahead depth. For $d = 1$, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1. 
The final P\textsubscript{d} is thresholded against prefetch threshold 
(T\textsubscript{p}) to reject the low confidence suggestions and 
then against a numerically bigger fill threshold (T\textsubscript{f}) to 
decide whether the prefetch should be sent to L2 Cache 
(high confidence prefetch) or Last Level Cache (low confidence prefetch).
The two thresholds were empirically set to 25 and 90 respectively, 
on the scale of 0 to 100.

@@

As was noted in Figure~\ref{Fig:Motivation}, an aggressive lookahead 
prefetching if done without any accuracy check, can harm the performance 
of the system. In this given case, the effective performance gained
by the original prefetcher is lost by almost 9\% by the aggressive version 
of the same prefetcher. Most modern prefetchers are getting increasingly 
aggressive as that enables them to capture the complex memory access patterns 
seen in pointer-chasing applications. Thus, we need a mechanism that is 
orthogonal to the underlying prefetching scheme and can be used
to prune out the useful prefetches from the useless ones.

>>>>

Kim {\em et. al.} proposed Signature Path Prefetcher (SPP)~\cite{SPP},
a confidence-based lookahead prefetcher.  SPP creates a signature
associated with a page address by compressing the history of
accesses. By correlating the signature with future likely delta
patterns, SPP learns both simple and complicated memory access
patterns quickly.  While the basic idea of perceptron based prefetch
filtering is applicable to any look-ahead prefetcher, we develop a
practical implementation of our proposed perceptron prefetch filter
using SPP as our baseline.  Here we describe the basic architecture of
SPP.

\noindent \textbf{Signature Table:} As shown on the right side of
figure \ref{fig:spp_update}, the Signature Table keeps track of 256
most recently accessed pages. It is meant to capture memory access
patterns within a page boundary. SPP indexes into an entry of the
Signature Table using the page number. For each entry corresponding to
a page, the table stores a `last block offset' and an `old
signature'. Last block offset is the block offset of the last memory
access of that given page. The block offset is calculated with respect
to the page boundary. The signature is a 12-bit compressed
representation of the past few memory accesses for that page. The
signature is calculated as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\;
(\,Delta\,)$$ Delta is the numerical difference between the block
offset of the current and the previous memory access. In case a
matching page entry is found, the stored signature retrieved and used
to index into the Pattern Table.  This process is illustrated in
figure \ref{fig:spp_update}.

\noindent \textbf{Pattern Table:} The Pattern Table, shown on the left
side of figure~\ref{fig:spp_update} is indexed by the signature
generated from the Signature Table.  Pattern Table holds predicted
delta patterns and their confidence estimates. Each entry indexed by
the signature holds up to 4 unique delta predictions.

@@

%% NOTE: Can we do away with this example? 
%% If it looks like we are writing too much about SPP
%% EDIT: Decided to reorganize and remove this for now.
%% 	Consider the case that incoming page number 10 with a block offset 3
%% 	finds a match in the ST. The retrieved pattern signature is 0x30 and
%% 	the Last Offset is 1. Since now there is the Last Block Offset (1),
%% 	Incoming Block Offset (3), Old Signature (0x30), and New Signature
%% 	calculated as per above equation (0x182), SPP can infer
%% 	non-speculatively that the given pattern of memory accesses (as
%% 	captured in the Old Signature) leads to the particular Delta. In
%% 	general, Delta is defined as the difference between the prefetch
%% 	suggestion block and the initial block which triggered the prefetch.
%% 	In this case, since SPP is in learning phase, it is defined as the
%% 	difference between the Incoming Block Offset and the Last Block Offset
%% 	(+2 in this case). This, in turn generates the new memory pattern
%% 	(New Signature). This newly learned signature pattern and the delta
%% 	is stored in the Pattern Table.

%  -- This looks like a detail that is confusing and not necessary to
%  understand SPP:
% \noindent \textbf{Prefetch Table:} The Prefetch Table is a 1024 entry
% direct-mapped table that keeps a record of last few entries
% prefetched.  This proves useful for updating the prefetcher state if a
% tracked prefetch leads to a demand hit or a cache
% eviction. \textit{Note:} The original paper refers to this as the
% Prefetch Filter but since in our context filter refers to the
% perceptron filter, we will be calling this structure as the Prefetch
% Table to avoid any confusion.

\noindent \textbf{Lookahead Prefetching:} On each trigger, SPP goes
down the program speculation path using its own prefetch suggestion.
Using current prefetch as a starting point, it re-accesses the Pattern
Table to generate further prefetches.  As illustrated in
figure~\ref{fig:spp_structure}, it repeats the cycle of accessing
Pattern Table and updating the signature based on highest confidence
prefetch from the last iteration.  The iteration counter on which SPP
manages to predict prefetch entries in the lookahead manner is
characterized as its `depth'. While doing so, SPP also keeps
compounding the confidence in each depth. Thus as depth increases,
overall confidence keeps decreasing.

\noindent \textbf{Confidence Tracking}: As shown in
figure~\ref{fig:spp_structure}, the Pattern Table keeps a count of
hits to each signature through a counter C\textsubscript{sig}. The
number of hits for a given delta per signature are tracked using a
counter C\textsubscript{delta}.  The confidence for a given delta is
approximated through C\textsubscript{d} = C\textsubscript{delta} /
C\textsubscript{sig}.  When SPP enters into a lookahead mode, the path
confidence P\textsubscript{d} is given as:
$$P\textsubscript{d} \;=\; \alpha  \;.\;  C\textsubscript{d}  \;.\;
P\textsubscript{d-1}$$ Here $\alpha$ represents the global accuracy,
calculated as the ratio of the number of prefetches which led to a
demand hit to the number of prefetches recommended in total. The range
of $\alpha$ is [0,1].  $D$ is the lookahead depth. For $d = 1$, when
SPP is in non-speculative mode, P\textsubscript{0} can be thought of
as 1.  The final P\textsubscript{d} is thresholded against prefetch
threshold (T\textsubscript{p}) to reject the low confidence
suggestions and then against a numerically bigger fill threshold
(T\textsubscript{f}) to decide whether the prefetch should be sent to
L2 Cache (high confidence prefetch) or Last Level Cache (low
confidence prefetch).  The two thresholds were empirically set to 25
and 90 respectively, on the scale of 0 to 100.

@@

As was noted in Figure~\ref{Fig:Motivation}, an aggressive lookahead
prefetching, if done without any accuracy check, can harm the
performance of the system. As the figure shows, aggressive lookahead
and its accompanied loss of accuracy degrades performance by almost
9\%.  This is despite a growing number of useful prefetches generated
by the prefetcher.  Thus, we need a mechanism that is orthogonal to
the underlying prefetching scheme and can be used to prune out the
useful prefetches from the useless ones.
